---
title: "Data Cleanup"
author: "Bartol Freškura"
date: "February 10, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

```{r, message=F}
# Loading packages
library(tidyverse)
library(VIM)
library(magrittr)
```

## Ratings

```{r}
ratings <- read.csv(file="data/goodbooks-10k/ratings.csv", stringsAsFactors = F,
                    colClasses = c("integer", "integer", "factor"))
```

```{r}
head(ratings)
str(ratings)
```
```{r}
# Check of missing data
sum(complete.cases(ratings)) == nrow(ratings)
```
There is no missing data!

```{r}
# Check for duplicate ratings
sum(duplicated(ratings[, 1:2])) == 0
```
No user has rated the same book more than once.

## Books

```{r}
books <- read.csv(file="data/goodbooks-10k/books.csv", stringsAsFactors = F, na.strings = "")
```

```{r}
head(books)
str(books)
```


```{r, fig.height=7, fig.width=10}
# Check for missing data
nrow(books) - sum(complete.cases(books))
aggr(books, prop=FALSE, numbers=TRUE)
lapply(books, function(x){sum(is.na(x))}) %>% as.data.frame()
```
There are 2140 rows with at least one missing column. Most of the missing values come from the 
*language_code* followed by the *isbn*, *isb13*, and *original_title*. Fortunetaly, all the 
variables which have missing data are either not crucial/interesting for the analysis or
don't have many missing values which would make them totally unuseful
(*original_publication_year* with 21 missing values).


```{r}
# Convert columns to more appropriate types
books$language_code = as.factor(books$language_code)
```

## Tags
```{r}
tags <- read.csv(file="data/goodbooks-10k/tags.csv", stringsAsFactors = F)
```

```{r}
head(tags)
str(tags)
```

```{r}
# Check of missing data
sum(complete.cases(tags)) == nrow(tags)
```
No missing data.

## Book Tags

```{r}
book_tags <- read.csv(file="data/goodbooks-10k/book_tags.csv", stringsAsFactors = F)
```

```{r}
head(book_tags)
str(book_tags)
```
```{r}
# Check of missing data
sum(complete.cases(book_tags)) == nrow(book_tags)
```
No missing data.


## To Read
```{r}
to_read <- read.csv(file="data/goodbooks-10k/to_read.csv", stringsAsFactors = F)
```

```{r}
head(to_read)
str(to_read)
```
```{r}
# Check of missing data
sum(complete.cases(to_read)) == nrow(to_read)
```
No missing data.

```{r}
# Check for duplicate ratings
sum(duplicated(to_read[, 1:2])) == 0
```
No duplicates are present.