---
title: "Exploratory Analysis"
author: "Bartol Freškura"
date: "February 10, 2018"
output:
    html_document:
        toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

```{r, message=F}
# Loading packages
library(tidyverse)
library(magrittr)
library(gridExtra)  
source('./scripts/data_preparation.R')
```

# Basic analysis

## Book tags
```{r}
head(book_tags)
glimpse(book_tags)
```
### Variables

* goodreads_book_id: Book ID extracted from the Goodreads
* tag_id: ID mapped to a specific tag description
* count: Number of tags for the given tag_id

```{r}
boxplot(book_tags$count)
summary(book_tags$count)
```
The minimum number of counts is -1 which doesn't seem right. Looking at the boxplot we can see that
the distribution is uneven with a lot of high count outliers. 

```{r}
sum(book_tags$count <= 0)
# Remove counts <= 0
book_tags <- filter(book_tags, count > 0)
```
There are 6 observations in which count variable is <= 0. These observations are useless for the 
analysis so we remove them.

## Books

```{r}
head(books)
glimpse(books)
```

### Variables

* book_id: Arbitrarily given ID
* goodreads_book_id: Book ID extracted from the Goodreads
* best_book_id: Book ID pointing to the most popular edition of a given book
* work_id: Book ID pointing to the book in the abstract sense.
* books_count: NA
* isbn: ISBN-10
* isbn13: ISBN-13
* authors: Book authors
* original_publication_year: Publication year (negative means B.C.)
* original_title: Original Title
* title: Title extracted from the Goodreads
* language_code: Language code in which the book was written I assume
* average_rating: Average rating extracted from the Goodreads
* ratings_count: Number of ratings for this version of the book
* work_ratings_count: Number of ratings for all versions of the book
* work_text_review_counts: Number of reviews for all versions of the book
* ratings_1: Number of ratings of value 1
* ratings_2: Number of ratings of value 2
* ratings_3: Number of ratings of value 3
* ratings_4: Number of ratings of value 4
* ratings_5: Number of ratings of value 5
* image_url: Book cover image URL
* small_image_url: Small book cover image URL

Let's do a quick analysis of variables which could be interesting

```{r}
ggplot(data=books, aes(x=language_code)) + geom_bar() + 
    theme(axis.text.x = element_text(angle=90, hjust=1))
```
Most of the books are written in some form of the English language which was expected. Other
languages are evenly distributed.

```{r, fig.width=10}
p1 <- ggplot(data=books, aes(x=original_publication_year)) + geom_density(na.rm = T)

p2 <- ggplot(data=filter(books, original_publication_year >= 1800), aes(x=original_publication_year)) + geom_density(na.rm = T)

p3 <- ggplot(data=filter(books, original_publication_year >= 1990), aes(x=original_publication_year)) + geom_density(na.rm = T)

grid.arrange(p1, p2, p3, ncol=2, nrow=2)
```
Publication year distribution is heavily skewed towards modern era as it can be seen on the
graphs above. Number of books starts to rise exponentially around 1900 and peaks around
year 2000. Zooming in further, the peak is seen at the year 2011.

```{r}
g <- ggplot(data=books, aes(x=average_rating, color="blue")) + geom_density(na.rm = T, color="blue")

normal <- rnorm(nrow(books), mean=mean(books$average_rating),
                sd=sd(books$average_rating))

g + geom_density(aes(x=normal), color="red")
```
Average ratings (blue) form a distribution very close to a normal one (red) with an unusual
"dent" around the 4.0 rating. 

```{r, fig.width=10}
p1 <- ggplot(data=filter(books,ratings_1 < 1500), aes(x=ratings_1)) + geom_histogram(na.rm = T, bins = 13)
p2 <- ggplot(data=filter(books,ratings_2 < 3000), aes(x=ratings_2)) + geom_histogram(na.rm = T, bins=13)
p3 <- ggplot(data=filter(books,ratings_3 < 10000), aes(x=ratings_3)) + geom_histogram(na.rm = T, bins=13)
p4 <- ggplot(data=filter(books,ratings_4 < 20000), aes(x=ratings_4)) + geom_histogram(na.rm = T, bins=13)
p5 <- ggplot(data=filter(books,ratings_5 < 20000), aes(x=ratings_5)) + geom_histogram(na.rm = T, bins=13)

grid.arrange(p1, p2, p3,p4, p5, ncol=2, nrow=3)
```
Ratings 4 and 5 have a quite similar distribution with most books having around 5000 ratings.
As we go to the lower ratings, a pattern emerges in which the number of ratings drops but the 
count increases. 

From these 5 graphs we can that an average rating from the previous graph can be attributed mostly
to ratings 3, 4, and 5 which skew the average rating towards higher rating.

```{r}
g <- ggplot(filter(books, work_ratings_count < 50000))
g <- g+ geom_density(aes(work_ratings_count), fill="blue", alpha=0.3)
g <- g+ geom_density(aes(work_text_reviews_count), fill="red",alpha=0.3)
g

median(books$work_ratings_count)
median(books$work_text_reviews_count)
```
One last thing to show is the ratio between the number of text reviews and ratings. We expect
the number of ratings to be higher than the number of text reviews but it is interesting to
see the ratio visually. Text reviews are packed in tight interval when compared to ratings.
This is also confirmed by the medians of both variables.


## Ratings

```{r}
head(ratings)
str(ratings)
```
### Variables

* User ID: Goodreads user ID
* book_id: *book_id* key variable from the *books* dataset
* rating: Rating (1-5) 

```{r}
ggplot(data=ratings, aes(rating)) + geom_bar(aes(fill=rating))
```
Another graph which confirmes the skew and the average rating from the former observations.
The cause for the skew is more obvious here due to the significantly smaller number of 1 and 2 ratings seen
on the graph.

## Tags

```{r}
head(tags)
glimpse(tags)
```

### Variables

* tag_id: Tag ID
* tag_name: Tag name



## To Read

```{r}
head(to_read)
glimpse(to_read)
```

# In-depth analysis

## What are the top 10 books by the number of text reviews, number of ratings, and the average rating

```{r, fig.width = 10}
n <- 10
top_10_text_review <- arrange(books, desc(work_text_reviews_count)) %>% head(n=n)
top_10_ratings <- arrange(books, desc(work_ratings_count)) %>% head(n=n)
top_10_avg_rating <- arrange(books, desc(average_rating)) %>% head(n=n)

ggplot(top_10_text_review, aes(x=reorder(title, -work_text_reviews_count), y=work_text_reviews_count, fill=title)) +
    geom_bar(stat = "identity") + labs(x="Title", y="Text Review Count") +
    theme(axis.text.x = element_text(size=0))

ggplot(top_10_ratings, aes(x=reorder(title, -ratings_count), y=ratings_count, fill=title)) +
    geom_bar(stat = "identity") + labs(x="Title", y="Ratings Count") +
    theme(axis.text.x = element_text(size=0))

# TODO graph is empty if the ylim is set to a smaller interval
ggplot(top_10_avg_rating, aes(x=reorder(title, -average_rating), y=average_rating, fill=title)) +
    geom_bar(stat = "identity") + labs(x="Title", y="Average rating") +
    theme(axis.text.x = element_text(size=0)) + ylim(0, 5)
```
It is more interesting to see which book come up top if we take a score which takes both
the number of ratings and the 

## Who are the best authors?
```{r, fig.width=10}
# Split authors so ever author is in one observation
by_author <- books %>% mutate(authors = strsplit(authors, ",")) %>% unnest(authors)

by_author <- group_by(by_author, authors) %>% summarise(avg_rating=mean(average_rating),
                                                    ratings_count=sum(ratings_count)) 

ggplot(top_n(by_author,n = 10, ratings_count), aes(x=reorder(authors, -ratings_count),
                                                    y=ratings_count, fill=authors)) +
    geom_bar(stat = "identity") + labs(x="Authors", y="Ratings Count") +
    theme(axis.text.x = element_text(size=0))

ggplot(top_n(by_author,n = 10, avg_rating), aes(x=reorder(authors, -avg_rating),
                                                    y=avg_rating, fill=authors)) +
    geom_bar(stat = "identity") + labs(x="Authors", y="Average rating") +
    theme(axis.text.x = element_text(size=0)) 

```

## Ratings count - Average ratings model

```{r, fig.width=10, fig.height=8}
alpha_even <- 0.5
alpha_ratings <- 0.75
alpha_count <- 1 - alpha_ratings

even <- transmute(books, title,  score = alpha_even*c(scale(average_rating)) +
                                     (1-alpha_even)*c(scale(work_ratings_count))) %>% top_n(n=10)

ratings_preffered <- transmute(books, title,  score = alpha_ratings*c(scale(average_rating)) +
                                     (1-alpha_ratings)*c(scale(work_ratings_count))) %>% top_n(n=10)

count_preffered <- transmute(books, title,  score = alpha_count*c(scale(average_rating)) +
                                     (1-alpha_count)*c(scale(work_ratings_count))) %>% top_n(n=10)

p1 <- ggplot(even, aes(x=reorder(title, -score), y=score, fill=title)) +
    geom_bar(stat = "identity") + labs(x="Title", y="Score", title="Alpha = 0.5") +
    theme(axis.text.x = element_text(size=0))
p2 <- ggplot(ratings_preffered, aes(x=reorder(title, -score), y=score, fill=title)) +
    geom_bar(stat = "identity") + labs(x="Title", y="Score", title="Alpha = 0.75") +
    theme(axis.text.x = element_text(size=0))
p3 <- ggplot(count_preffered, aes(x=reorder(title, -score), y=score, fill=title)) +
    geom_bar(stat = "identity") + labs(x="Title", y="Score", title="Alpha = 0.25") +
    theme(axis.text.x = element_text(size=0))

grid.arrange(p1, p2, p3, ncol=1, nrow=3)
```

Three graphs above show us different rankings when we apply a simple model which takes
standardised average ratings and rating counts to calculate the book score. Alpha parameter
represents the weight given to the average rating variable.

## Tags analysis

```{r}
tags_grouped <- inner_join(book_tags, tags, by=c('tag_id'='tag_id')) %>% group_by(tag_name, tag_id) %>%
    summarise(count=sum(count)) %>% arrange(desc(count))

head(tags_grouped, n = 200)
```


```{r}
ggplot(head(tags_grouped, n=50), aes(x=reorder(tag_name, -count), y=count)) +
    geom_bar(stat = "identity") + labs(x="Tag name", y="Count (log10)", title="50 most popular tags") + 
    theme(axis.text.x = element_text(angle = 90)) + 
    scale_y_log10()
```
Grouping tags by the tag count gives us a quite complex result to analyze. *to-read* and
*currently-reading* are the default categories that Goodreads offer so it was expected to find
them at the top. I was expecting that the tags popularity will perfectly correspond to book genres but
that is not the case. Tags like *my-books*, *to-buy*, or *library* make it much harder to
filter and analyze just the book genres so I will split the analysis into two parts. The first
part will consider only book genres and the second part will pick non-genre tags that seem interesting.

One more issue are the duplicate tags, e.g. *ya* and *young-adult*. If there are such tags, I 
will consider only the most popular one. I chose this approach because the grouping can become
burdensome as some of the duplicates can appear very far down the popularity list and will
take a lot of time to properly group them.

### Book genres analysis
As there are too many genres to make a complete analysis, I will only focus on the most
popular and the most interesting (in my opinion :) ) ones. 

```{r}
# Pick genre IDs
pop_ids <- c(11305, 33114, 7457, 26138, 20939, 21689, 26837, 22973, 14821, 6953, 13561, 30358,
             1691, 14552, 15048, 8517, 19733, 4605, 19052, 23471, 26816, 5951,
             32130, 25647, 27095, 23931, 24526)

filter(tags_grouped, tag_id %in% pop_ids)
```

```{r}
popular_genres <- filter(book_tags, tag_id %in% pop_ids) %>% inner_join(books, by=c('goodreads_book_id' = 'goodreads_book_id'))
popular_genres <- inner_join(popular_genres, tags, by="tag_id")

# Even rating and rating count weight
popular_genres <- mutate(popular_genres, score=(c(scale(average_rating)) +
                                                    c(scale(work_ratings_count))))
head(popular_genres)
```

```{r, fig.width=10, fig.height=15}
p1 <- ggplot(popular_genres, aes(x=tag_name, y=average_rating)) +
    geom_boxplot(outlier.alpha = 0.3) + labs(x="Genre", y="Average rating") + 
    theme(axis.text.x = element_text(angle = 45))

p2 <- ggplot(popular_genres, aes(x=tag_name, y=work_ratings_count)) +
    geom_boxplot(outlier.alpha = 0.3) + labs(x="Genre", y="Number of ratings (log10)") + 
    theme(axis.text.x = element_text(angle = 45)) +
    scale_y_log10()

p3 <- ggplot(popular_genres, aes(x=tag_name, y=work_text_reviews_count)) +
    geom_boxplot(outlier.alpha = 0.3) + labs(x="Genre", y="Number of text ratings (log10)") + 
    theme(axis.text.x = element_text(angle = 45))  + scale_y_log10()
    
grid.arrange(p1, p2, p3, nrow=3)
```
* Ratings have more negative outliers
* manga and graphic novels higher scores but less ratings
* Ratings have more positive outliers (no negative ones)
* Textual reviews have an even number of positive/negative outliers

```{r}
draw_graphs <- function(data, title){
    P <- list()
    for(tag in unique(data$tag_name)){
        p <- ggplot(filter(data, tag_name==tag) , aes(x=reorder(title, -score), y=score)) +
            geom_bar(stat = "identity", aes(fill=title)) + labs(x="", y="Score", title=paste("Top", tag)) +
            theme(axis.text.x = element_text(size=8, angle = 40, hjust=1)) +
            guides(fill=FALSE) + theme(plot.margin = unit(c(0.3,0.1,0,3,0.1), "cm"))
        P <- c(P, list(p))
    }
    
    Plots <- list(plots=P, num=length(P))
    do.call(grid.arrange, c(Plots$plots, ncol=5, nrow=(Plots$num) %/% 5 + 1, top=title))
}
```

Here are the plots of top 15 books in each genre.
```{r, fig.width=40, fig.height=40}
top_popular_genres <- filter(popular_genres, c(scale(count)) > -1) %>%
    group_by(tag_name) %>% top_n(n=15, wt=score)

draw_graphs(top_popular_genres, "Top 15 books by the genre")
```

#### Book scores with regards to their "popularity bucket"

As we calcuate the score using the average rating and number of ratings, it would be
interesting to see top books if split them by their number of ratings which I will
call popularity from now on.

```{r}
# Bucket 1 has the most popular, 7 the least popular books
n <- 7
popular_genres$popularity <- ntile(desc(popular_genres$work_ratings_count), n)
```

```{r, fig.width=35, fig.height=40}
for(i in 2:n){
    bucket <- filter(popular_genres, popularity==i)
    bucket_top <- filter(bucket, c(scale(count)) > -1) %>%
        group_by(tag_name) %>% top_n(n=15, wt=score)
    draw_graphs(bucket_top, title=paste("Top 15 books by the genre in the popularity bucket", i))
}
```

 
### Other tags analysis
```{r}
# Pick interesting IDs
pop_ids <- c(11557, 5207, 17213, 4949, 3389, 24960, 26735, 1416)

filter(tags_grouped, tag_id %in% pop_ids)
```
```{r}
popular_other <- filter(book_tags, tag_id %in% pop_ids) %>% inner_join(books, by=c('goodreads_book_id' = 'goodreads_book_id'))
popular_other <- inner_join(popular_other, tags, by="tag_id")

# Even rating and rating count weight
popular_other <- mutate(popular_other, score=(c(scale(average_rating)) +
                                                    c(scale(work_ratings_count))))
head(popular_other)
```

```{r, fig.width=10, fig.height=15}
p1 <- ggplot(popular_other, aes(x=tag_name, y=average_rating)) +
    geom_boxplot(outlier.alpha = 0.3) + labs(x="Genre", y="Average rating") + 
    theme(axis.text.x = element_text(angle = 45))

p2 <- ggplot(popular_other, aes(x=tag_name, y=work_ratings_count)) +
    geom_boxplot(outlier.alpha = 0.3) + labs(x="Genre", y="Number of ratings (log10)") + 
    theme(axis.text.x = element_text(angle = 45)) +
    scale_y_log10()

p3 <- ggplot(popular_other, aes(x=tag_name, y=work_text_reviews_count)) +
    geom_boxplot(outlier.alpha = 0.3) + labs(x="Genre", y="Number of text ratings (log10)") + 
    theme(axis.text.x = element_text(angle = 45))  + scale_y_log10()
    
grid.arrange(p1, p2, p3, nrow=3)
```


```{r, fig.width=25, fig.height=13}
top_popular_other <- filter(popular_other, c(scale(count)) > -1) %>%
    group_by(tag_name) %>% top_n(n=15, wt=score)

draw_graphs(top_popular_other, "Top 15 books by the other tags")
```

#### Book scores with regards to their "popularity bucket"

```{r}
n <- 7
# Bucket 1 has the most popular, 7 the least popular books
popular_other$popularity <- ntile(desc(popular_other$work_ratings_count), n)
```

```{r, fig.width=25, fig.height=13}
for(i in 2:n){
    bucket <- filter(popular_other, popularity==i)
    bucket_top <- filter(bucket, c(scale(count)) > -1) %>%
        group_by(tag_name) %>% top_n(n=15, wt=score)
    draw_graphs(bucket_top, title=paste("Top 15 books by the other tags in the popularity bucket", i))
}
```


