---
title: "Exploratory Analysis"
author: "Bartol Freškura"
date: "February 10, 2018"
output:
    html_document:
        toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

```{r, message=F}
# Loading packages
library(tidyverse)
library(magrittr)
library(gridExtra)  
source('./scripts/data_preparation.R')
```

# Basic analysis

## Book tags
```{r}
head(book_tags)
glimpse(book_tags)
```
### Variables

* goodreads_book_id: Book ID extracted from the Goodreads
* tag_id: ID mapped to a specific tag description
* count: Number of tags for the given tag_id

```{r}
boxplot(book_tags$count)
summary(book_tags$count)
```
The minimum number of counts is -1 which doesn't seem right. Looking at the boxplot we can see that
the distribution is uneven with a lot of high count outliers. 

```{r}
sum(book_tags$count <= 0)
# Remove counts <= 0
book_tags <- filter(book_tags, count > 0)
```
There are 6 observations in which count variable is <= 0. These observations are useless for the 
analysis so we remove them.

## Books

```{r}
head(books)
glimpse(books)
```

### Variables

* book_id: Arbitrarily given ID
* goodreads_book_id: Book ID extracted from the Goodreads
* best_book_id: Book ID pointing to the most popular edition of a given book
* work_id: Book ID pointing to the book in the abstract sense.
* books_count: NA
* isbn: ISBN-10
* isbn13: ISBN-13
* authors: Book authors
* original_publication_year: Publication year (negative means B.C.)
* original_title: Original Title
* title: Title extracted from the Goodreads
* language_code: Language code in which the book was written I assume
* average_rating: Average rating extracted from the Goodreads
* ratings_count: Number of ratings for this version of the book
* work_ratings_count: Number of ratings for all versions of the book
* work_text_review_counts: Number of reviews for all versions of the book
* ratings_1: Number of ratings of value 1
* ratings_2: Number of ratings of value 2
* ratings_3: Number of ratings of value 3
* ratings_4: Number of ratings of value 4
* ratings_5: Number of ratings of value 5
* image_url: Book cover image URL
* small_image_url: Small book cover image URL

Let's do a quick analysis of variables which could be interesting

```{r}
ggplot(data=books, aes(x=language_code)) + geom_bar() + 
    theme(axis.text.x = element_text(angle=90, hjust=1))
```
Most of the books are written in some form of the English language which was expected. Other
languages are evenly distributed.

```{r, fig.width=10}
p1 <- ggplot(data=books, aes(x=original_publication_year)) + geom_density(na.rm = T)

p2 <- ggplot(data=filter(books, original_publication_year >= 1800), aes(x=original_publication_year)) + geom_density(na.rm = T)

p3 <- ggplot(data=filter(books, original_publication_year >= 1990), aes(x=original_publication_year)) + geom_density(na.rm = T)

grid.arrange(p1, p2, p3, ncol=2, nrow=2)
```
Publication year distribution is heavily skewed towards modern era as it can be seen on the
graphs above. Number of books starts to rise exponentially around 1900 and peaks around
year 2000. Zooming in further, the peak is seen at the year 2011.

```{r}
g <- ggplot(data=books, aes(x=average_rating, color="blue")) + geom_density(na.rm = T, color="blue")

normal <- rnorm(nrow(books), mean=mean(books$average_rating),
                sd=sd(books$average_rating))

g + geom_density(aes(x=normal), color="red")
```
Average ratings (blue) form a distribution very close to a normal one (red) with an unusual
"dent" around the 4.0 rating. 

```{r, fig.width=10}
p1 <- ggplot(data=filter(books,ratings_1 < 1500), aes(x=ratings_1)) + geom_histogram(na.rm = T, bins = 13)
p2 <- ggplot(data=filter(books,ratings_2 < 3000), aes(x=ratings_2)) + geom_histogram(na.rm = T, bins=13)
p3 <- ggplot(data=filter(books,ratings_3 < 10000), aes(x=ratings_3)) + geom_histogram(na.rm = T, bins=13)
p4 <- ggplot(data=filter(books,ratings_4 < 20000), aes(x=ratings_4)) + geom_histogram(na.rm = T, bins=13)
p5 <- ggplot(data=filter(books,ratings_5 < 20000), aes(x=ratings_5)) + geom_histogram(na.rm = T, bins=13)

grid.arrange(p1, p2, p3,p4, p5, ncol=2, nrow=3)
```
Ratings 4 and 5 have a quite similar distribution with most books having around 5000 ratings.
As we go to the lower ratings, a pattern emerges in which the number of ratings drops but the 
count increases. 

From these 5 graphs we can that an average rating from the previous graph can be attributed mostly
to ratings 3, 4, and 5 which skew the average rating towards higher rating.

```{r}
g <- ggplot(filter(books, work_ratings_count < 50000))
g <- g+ geom_density(aes(work_ratings_count), fill="blue", alpha=0.3)
g <- g+ geom_density(aes(work_text_reviews_count), fill="red",alpha=0.3)
g

median(books$work_ratings_count)
median(books$work_text_reviews_count)
```
One last thing to show is the ratio between the number of text reviews and ratings. We expect
the number of ratings to be higher than the number of text reviews but it is interesting to
see the ratio visually. Text reviews are packed in tight interval when compared to ratings.
This is also confirmed by the medians of both variables.


## Ratings

```{r}
head(ratings)
str(ratings)
```
### Variables

* User ID: Goodreads user ID
* book_id: *book_id* key variable from the *books* dataset
* rating: Rating (1-5) 

```{r}
ggplot(data=ratings, aes(rating)) + geom_bar(aes(fill=rating))
```
Another graph which confirmes the skew and the average rating from the former observations.
The cause for the skew is more obvious here due to the significantly smaller number of 1 and 2 ratings seen
on the graph.

## Tags

```{r}
head(tags)
glimpse(tags)
```

### Variables

* tag_id: Tag ID
* tag_name: Tag name



## To Read

```{r}
head(to_read)
glimpse(to_read)
```
